Using structural knowledge to correct RNA pyrosequencing errors in non-coding RNAs

### 2013-05-23 JCB/methods_JCB.tex
fd

The \emph{Inside} function $\Z{\Struct}{m}{x,y}$ is simply the partition function, i.e. the sum of Boltzmann factors over all sequences for a substructure $\Struct$ (implicitly attached to an interval $[i,j]$ of the sequence), featuring $m$ mutations/errors compared to $s$, and having flanking nucleotides $x$ and $y$. 
Such terms can be computed recursively, using the following equation for the initial case:
\begin{equation}
	\forall x,y\in B\times \B, m\in [0,M]:\, \Z{\varepsilon}{m}{x,y}=\left\{
	\begin{array}{ll}
		1 &\text{ If } m = 0\\
		0 &\text{ Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, either there is no sequence at distance $m>0$ of the empty sequence, or the only allowed sequence is the empty sequence ($m=0$), having energy $0$. Since the contributions only depend on base pairs, they do not appear in the initial conditions.

@@

The \emph{Outside} function, $\Y{\Struct}{m}{x,y}$ is the partition function considering only the 
contributions of subsequences excluding a given structure/interval $\Struct$, occupying the open interval $]i,j[$ in the sequence, at Hamming distance exactly $m$ to the initial sequence $s$, and assuming that nucleotides $x$ and $y$ were previously chosen for $i+1$ and $j-1$, the outermost portions of the excluded structure.
\TODOYann{Reformat outside Figure as in Incarnation.}
The associated terms $\Y{\Struct}{m}{x,y}$ can then be computed recursively, initially considering the case of any prefix $\Struct'$ of the complete structure:
\begin{equation}
	\forall x,y\in \B\times \B, \forall \Struct'\text{ s.t. }\Struct=\Struct'.\Struct'', m\in [0,M]:\, \Y{\Struct'}{m}{x,y}=\Z{\Struct''}{m}{y,z}
\label{eq:Y_in}
\end{equation}
where $z\in\B$ can be any nucleotide, and provably does not affect further computations. 
In other words, the sequences explored by an outside computation, excluding a prefix of $\Struct$, are exactly the sequences generated on the corresponded suffix. This set of sequence is also the inside term on the suffix structure.
It is also worth pointing out that $\Y{\Struct}{m}{x,y}= \Z{\varepsilon}{m}{y,X}=\mathbbm{1}_{m=0}$.

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires the precomputation and storage of the inside contributions.\label{fig:outside}}
\end{figure}
In general, the main recurrence below works by extending the excluded structure $\Struct$ (covering the $]i,j[$ interval) in the leftward direction. As shown in Figure~\ref{fig:outside}, four cases must be considered, depending on the base-pairing status and directionality of the next considered position:
\begin{itemize}
\item {\bf Case 1: Unpaired position.} When the innermost leftward base is unpaired, any base $a\in\B$ may be chosen for this position, and the rest of the sequence is generated recursively. The number of remaining mutations may be decreased by the choice of $a$, leading to the following recurrence:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'\in \B,\\ \Kron_{a,a'}\le m}}
    \Y{\ub\Struct}{m- \Kron_{a,a'}}{a',y}
\label{eq:Y_rec_A}
\end{equation}
\item {\bf Case 2: Stacking base-pair.} If both ends of the excluded structure are paired together, and are nested within another base-pair in the remaining structure, then an additional contribution for stacking has to be considered. The outside terms are then computed by simulating any pair of base-pairing nucleotides for $i,j$, and by proceeding recursively on the remaining portion, as follows:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 e^{\frac{-\EBP{i,j}{xy}{a'b'}}{RT}}\cdot
    \Y{\op\Struct\cp}{m- \Kron_{ab,a'b'}}{a',b'} 
\label{eq:Y_rec_B}
\end{equation}
\item {\bf Case 3: Next position paired rightward, in the absence of stacking pair.} In this case, the innermost leftward position $i$ is paired to the right at some position $k$ . 
Let us assume that its partner resides outside the excluded structure $\Struct$ (This assumption is provably without loss of generality, and directly follows from the fact that $\Struct$ is well-parenthesized). 
The structure within the base-pair may be described by the expression $\op\Struct \Struct'\cp$, where $\Struct'$ may possibly be empty (except if $\Struct$ is enclosed within corresponding brackets, to avoid stacking pairs). Therefore, any sequence considered by the outside computation consists in three independent parts: two nucleotides for the paired positions, a sequence for the region excluding $\Struct''$, and a sequence for $\Struct'$. It follows that:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
  		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct \Struct'\cp}{m- \Kron_{ab,a'b'} - m'}{a',b'}
     \cdot\Z{\Struct'}{m'}{y,b'} .
\label{eq:Y_rec_C}
\end{equation}
\item {\bf Case 4: Next position paired leftward.} If the innermost, leftward, position $i$ is paired to some position $k<i$, delimiting a substructure $\Struct'$, then any sequence considered by the outside computation consists in three parts: two nucleotides $a$ and $b$, a sequence for $\Struct'$, and a sequence for the region excluding $\op\Struct'\cp\Struct$. Consequently, one has:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   	 e^{\frac{-\EBP{k,i}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct'\cp\Struct}{m- \Kron_{ab,a'b'} - m'}{a',b}
     \cdot\Z{\Struct'}{m'}{a',b'}
\label{eq:Y_rec_D}
\end{equation}
\end{itemize}

>>>>

The \emph{Inside} function $\Z{\Struct}{m}{x,y}$ is simply the partition function, i.e. the sum of Boltzmann factors over all sequences for a substructure $\Struct$ (implicitly attached to an interval $[i,j]$ of the sequence), featuring $m$ mutations/errors compared to $s$, and having flanking nucleotides $x$ and $y$. 
Such terms can be computed recursively, using the following equation for the initial case:
\begin{equation}
	\forall x,y\in \B\times \B, m\in [0,M]:\, \Z{\varepsilon}{m}{x,y}=\left\{
	\begin{array}{ll}
		1 &\text{ If } m = 0\\
		0 &\text{ Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, either there is no sequence at distance $m>0$ of the empty sequence, or the only allowed sequence is the empty sequence ($m=0$), having energy $0$. Since the contributions only depend on base pairs, they do not appear in the initial conditions.

@@

The \emph{Outside} function, $\Y{\Struct}{m}{x,y}$ is the partition function considering only the 
contributions of subsequences excluding a given structure/interval $\Struct$, occupying the open interval $]i,j[$ in the sequence, at Hamming distance exactly $m$ to the initial sequence $s$, and assuming that nucleotides $x$ and $y$ were previously chosen for $i+1$ and $j-1$, the outermost portions of the excluded structure.
The associated terms $\Y{\Struct}{m}{x,y}$ can then be computed recursively, initially considering the case of any prefix $\Struct'$ of the complete structure:
\begin{equation}
	\forall x,y\in \B\times \B, \forall \Struct'\text{ s.t. }\Struct=\Struct'.\Struct'', m\in [0,M]:\, \Y{\Struct'}{m}{x,y}=\Z{\Struct''}{m}{y,z}
\label{eq:Y_in}
\end{equation}
where $z\in\B$ can be any nucleotide, and provably does not affect further computations. 
In other words, the sequences explored by an outside computation, excluding a prefix of $\Struct$, are exactly the sequences generated on the corresponded suffix. This set of sequence is also the inside term on the suffix structure.
It is also worth pointing out that $\Y{\Struct}{m}{x,y}= \Z{\varepsilon}{m}{y,X}=\mathbbm{1}_{m=0}$.

\begin{figure}[t]\centering
\includegraphics[scale=\ScaleDP]{FigDPOutsideWrapper}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires the precomputation and storage of the inside contributions.\label{fig:outside}}
\end{figure}
In general, the main recurrence below works by extending the excluded structure $\Struct$ (covering the $]i,j[$ interval) in the leftward direction. As shown in Figure~\ref{fig:outside}, four cases must be considered, depending on the base-pairing status and directionality of the next considered position:
\begin{itemize}
\item {\bf Case 1: Unpaired position.} When the innermost leftward base is unpaired, any base $a\in\B$ may be chosen for this position, and the rest of the sequence is generated recursively. The number of remaining mutations may be decreased by the choice of $a$, leading to the following recurrence:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'\in \B,\\ \Kron_{a,a'}\le m}}
    \Y{\ub\Struct}{m- \Kron_{a,a'}}{a',y}
\label{eq:Y_rec_A}
\end{equation}
\item {\bf Case 2: Stacking base-pair.} If both ends of the excluded structure are paired together, and are nested within another base-pair in the remaining structure, then an additional contribution, stemming from a stacking pairs energy, has to be considered. The outside terms are then computed by simulating any pair of base-pairing nucleotides for $i,j$, and by proceeding recursively on the remaining portion, as follows:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 e^{\frac{-\EBP{i,j}{xy}{a'b'}}{RT}}\cdot
    \Y{\op\Struct\cp}{m- \Kron_{ab,a'b'}}{a',b'} 
\label{eq:Y_rec_B}
\end{equation}
\item {\bf Case 3: Next position paired rightward, in the absence of stacking pair.} In this case, the innermost leftward position $i$ is paired to the right at some position $k$ . 
Let us assume that its partner resides outside the excluded structure $\Struct$ (This assumption is provably without loss of generality, and directly follows from the fact that $\Struct$ is well-parenthesized). 
The structure within the base-pair may be described by the expression $\op\Struct \Struct'\cp$, where $\Struct'$ may possibly be empty (except if $\Struct$ is enclosed within corresponding brackets, to avoid stacking pairs). Therefore, any sequence considered by the outside computation consists in three independent parts: two nucleotides for the paired positions, a sequence for the region excluding $\Struct''$, and a sequence for $\Struct'$. It follows that:
\begin{equation}
	\Y{\Struct}{m}{x,y} = \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
  		 e^{\frac{-\EBP{i,k}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct \Struct'\cp}{m- \Kron_{ab,a'b'} - m'}{a',b'}
     \cdot\Z{\Struct'}{m'}{y,b'} .
\label{eq:Y_rec_C}
\end{equation}
\item {\bf Case 4: Next position paired leftward.} If the innermost, leftward, position $i$ is paired to some position $k<i$, delimiting a substructure $\Struct'$, then any sequence considered by the outside computation consists in three parts: two nucleotides $a$ and $b$, a sequence for $\Struct'$, and a sequence for the region excluding $\op\Struct'\cp\Struct$. Consequently, one has:
\begin{equation}
	\Y{\Struct}{m}{x,y} = 
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{ab,a'b'}\le m}}
		 \sum_{m'=0}^{m-\Kron_{ab,a'b'}}
   	 e^{\frac{-\EBP{k,i}{\varnothing}{a'b'}}{RT}}
		 \cdot\Y{\op\Struct'\cp\Struct}{m- \Kron_{ab,a'b'} - m'}{a',b}
     \cdot\Z{\Struct'}{m'}{a',b'}
\label{eq:Y_rec_D}
\end{equation}
\end{itemize}
